Primal-dual algorithm for computing Nash equilibria in sequential games with imcomplete information

### 2015-01-08 nips2015.tex
typos
WIP

\paragraph{Nash equilibrium problem in sequence-form}
In \cite{von1996efficient, vonequilibrium}, it was established that for sequential two-person zero-sum games with imcomplete information and perfect recall, there exists sparse matrices $A \in \mathbb{R}^{n_1 \times n_2}, E_1 \in \mathbb{R}^{l_1 \times n_1}, E_2 \in \mathbb{R}^{l_2 \times n_2}$, and vectors $e_1 \in \mathbb{R}^{l_1}, e_2 \in \mathbb{R}^{l_2}$ such that $n_1$, $n_2$, $l_1$, and $l_2$ are all linear in the size of the game tree (number of states in the game) and such that Nash equilibria correspond to points $(x, y)$ which solve the primal LCP (Linear Constrained Programming) problem
\begin{equation}
  \begin{aligned}
    & \underset{y,p}{\text{minimize}}
    & & e_1^Tp \\
    & \text{subject to}
    & & -Ay + E_1^Tp \geq 0,\\
    &&& E_2y = e_2,\\
    &&& y \ge 0
  \end{aligned}
  \label{eq:primal_pb}
\end{equation}

@@

with duality gap given by
\begin{equation}
  \mathcal{G}(p, q) := e_1^Tp - (-e_2^Tq) = e_1^Tp + e_2^Tq
  \label{eq:dgap}
\end{equation}

@@

Of course by \textit{weak duality} in problems \eqref{eq:primal_pb} and \eqref{eq:dual_pb},, it holds that
\begin{equation}
  \mathcal{G}(p, q) \geq 0
\end{equation}
for any fixed pair $(x, y)$ of realization plans and admissible dual pair $(p,q)$ in problems \eqref{eq:primal_pb} and \eqref{eq:dual_pb}.

It was also shown (for example, Theorem 3.14 of \cite{vonequilibrium}) that a $(x, y)$ is a solution to the primal-dual problem \eqref{eq:primal_pb} and \eqref{eq:dual_pb} iff there exists vectors $p$ and $q$ such that
\begin{equation}
  \begin{split}
    E_1x = e_1,\text{ }x \geq 0,\hspace{1em} E_2y = e_2,\text{ }y \geq 0\\
    -Ay + E_1^Tp \geq 0,\hspace{1em}-A^Tx + E_2^Tq \leq 0
    \end{split}
\end{equation}

Moreover, in this case \textit{strong duality} holds and the value of the game equals $e_1^Tp = -e_2^Tq$,
In this case, the duality gap defined in \eqref{eq:dgap} vanishes.

@@

\paragraph{Outline of article}
The goal of this article is to device a simple algorithm which produces a Nash $\epsilon$-equilibrium after $\mathcal{O}(1/\epsilon)$ iterations%We first briefly introduce the sequence-form representation, and contrast it to the strategic-form representation. Then we state the Nash equilibrium problem for the sequence-form representation. Then in
In section \ref{sec:related_work}, we give a brief overview of existing methods for solving \eqref{eq:approx_pb}. In section \ref{sec:algo} we derive the proposed algorithm for solving this problem and prove its $\mathcal{O}(1/\epsilon)$ convergence rate.  This is followed by a demonstration on Kuhn Poker in section \ref{sec:kuhn}. In section \ref{sec:algo_gen}, we extend  the algorithm to problems with more general strategy profiles, and exhibit thesame convergence properties.

@@

\section{Related work}
\label{sec:related_work}
EGT used in \cite{hoda2010smoothing}...

\section{Useful convex analysis}
\label{sec:notation}
We will need the following notations and definitions in the sequel. Given positive integers $m$ and $n$, $\mathbb{R}^{m \times n}$ denotes
the space of all $m$-by-$n$ real matrices. $0_{m,n}$ denotes the $m$-by-$n$ matrix of zeros and $1_{m,n}$ denotes the $m$-by-$n$ matrix of ones.
%% $\mathbb{R}^n_+$ := $\{x \in \mathbb{R}^n|x_k \ge 0 \text{ }  \forall j\}$ is the $n$-dimensional \textit{nonnegative orthant}.

@@

Note that $i_{C \cap D} = i_C + i_D$. The \textit{euclidean projection operator} onto $C$, denoted $\Pi_C$ is the function
$\Pi: \mathbb{R}^n \mapsto C$, which maps a point $x \in \mathbb{R}^n$ to the (necessarily unique) point $\Pi_C(x)$ of $C$ which is closed to $x$. Precisely,
\begin{equation}
  \Pi_C(x) := \underset{c \in C}{argimin}\text{ }\|c - x\|^2
\end{equation}
For example, the metric projections onto the nonnegative orthant and the closed unit-disk $\bar{\Delta}$
% := \{x|\|x\| \le 1\}$
are $\Pi_{\mathbb{R}^n_+}(x) \equiv (x)_+$ and $\Pi_{\bar{\Delta}}(z) \equiv \dfrac{z}{max(\|z\|, 1)}$ respectively.

Let $f : \mathbb{R}^n \rightarrow [-\infty, +\infty]$ be a \textit{proper convex lower semi-continous function}
(\textit{p.c.l.s.c} for short). The \textit{convex conjugate} of $f$ is the function $f^*: \mathbb{R}^n \rightarrow [-\infty, +\infty]$ is defined by
\begin{equation}
  f^*(x) := \underset{z \in \mathbb{R}^n}{\text{max}}\text{ }z^Tx - f(x)
\end{equation}

@@

\section{The proposed algorithm}
\label{sec:algo}
We now derive our primal-dual algorithm for solving the Nash equilibrium problem \eqref{eq:primal_pb}. The algorithm is derived from the general primal-dual Algorithm 39 proposed in \cite{chambolle2010}. Though this scheme has recently gained considerable popularity in the signal processing community
%\cite{gramfort-etal:2013a}
, to the best of our knowledge, this is the first time it is being applied to compute Nash equilibria.

It should be noted that the difficulty of applying the general primal-dual Algorithm 39 proposed in \cite{chambolle2010} directly to problem \eqref{eq:primal_pb} lies in the difficulty of effectively computing the metric projections $\Pi_{Q_k}$. Theorem \ref{thm:pd} provides a trick to effectively evade this difficulty.

\begin{theorem}
  The Nash equilibrium LCP problems \eqref{eq:primal_pb} and \eqref{eq:dual_pb} can be re-written in the saddle-point form

@@

Now,
\begin{equation}
  \begin{split}
    H^*\left(K\begin{bmatrix}y\\p\end{bmatrix}\right) := \underset{x,q}{\text{max}}\text{ }\begin{bmatrix}x\\q\end{bmatrix}^TK\begin{bmatrix}y\\p\end{bmatrix} - H(x, q) &= \underset{x, q}{\text{max}}\text{ }x^T(Ay - E_1^Tp) - i_{\mathbb{R}^{n_1}_+}(x) + q^TE_2y - f^Tq\\
      &= \underset{x \geq 0}{\text{max}}\text{ }x^T(Ay - E_1^Tp) + \underset{q}{\text{max}}\text{ }q^TE_2y - f^Tq\\
      &= i_{\mathbb{R}^{n_1}_+}(-Ay + E_1^Tp) + i_{\{0\}}(E_2y - e_2)
  \end{split}
  \label{eq:b}
\end{equation}
Combining equations \eqref{eq:a} and \eqref{eq:b}, it follows that the LCP problem \eqref{eq:primal_pb} can be written as 
\begin{equation}
  \underset{y,p}{\text{minimize}}\text{ }G(y,p) + H^*\left(K\begin{bmatrix}y\\p\end{bmatrix}\right)
\end{equation}

@@

But this two last equations are simply the primal and dual formulations of the saddle-point problem \eqref{eq:unconstrained_pb}, and we are done.

@@

\begin{algorithm}[H]
  \caption{$\mathcal{O}(1/\epsilon)$ Primal-dual algorithm for finding a Nash $\epsilon$-equilibrium for a sequential two-person zero-sum game with imcomplete information and perfect recall}
  \KwIn{sequence-form specification of the game: $(A, E_1, E_2, e_1, e_2)$, where $A \in \mathbb{R}^{n_1 \times n_2}$,
  $E_1 \in \mathbb{R}^{l_1 \times n_1}$, $E_2 \in \mathbb{R}^{l_2 \times n_2}$, $e_1 \in \mathbb{R}^{l_1}$, $e_2 \in \mathbb{R}^{l_2}$}
  \KwOut{Nash $\epsilon-$equilibrium pair $(x^{(k)}$, $y^{(k)})$ of realization plans}
  \textbf{Precompute}: $\|K\|^2$, where $K$ is constructed as in equations \eqref{eq:unconstrained_pb}. $\|K\|^2$ can be computed via a \textit{power iteration} on $K^TK$, for example.\\
  \textbf{Initialize}:
  $x^{(0)} \in \mathbb{R}^{n_1}$; $\tilde{y^{(0)}}, y^{(0)} \in \mathbb{R}^{n_2}$; $p \in \mathbb{R}^{l_1}$; $q^{(0)} \in \mathbb{R}^{l_2}$; 
  $\tau, \sigma > 0 \text{ s.t. }\tau\sigma \|K\|^2 < 1$ (for example take $\tau = \sigma = .99/\|K\|$); $k = 0$.\\
  \While{
%$\dfrac{\| x^{(k+1)} - x^{(k)}\|^2 + \|v^{(k+1)}- v^{(k)}\|^2}{\sigma} + \dfrac{\|y^{(k+1)}- y^{(k)}\|^2 + \|u^{(k+1)}- u^{(k)}\|^2}{\tau} < \epsilon$
$|e_1^Tp^{(k)} - e_2^Tq^{(k)}| \geq \epsilon$}{
    \begin{eqnarray*}
      x^{(k+1)} &\leftarrow& \left(x^{(k)} + \tau \left(A\tilde{y}^{(k)} - E_1^T\tilde{p}^{(k)}\right)\right)_+\\
      q^{(k+1)} &\leftarrow& q^{(k)} + \tau \left(E_2\tilde{y}^{(k)} - e_2\right)\\
      y^{(k+1)} &\leftarrow& \left(y^{(k)} - \sigma \left(A^Tx^{(k + 1)} + E_2^Tq^{(k + 1)}\right)\right)_+\\
      p^{(k+1)} &\leftarrow& p^{(k)} - \sigma \left(e_1 - E_1x^{(k+1)}\right)\\
      \tilde{y}^{(k+1)} &\leftarrow& 2y^{(k+1)} - y^{(k)}\\
      \tilde{q}^{(k+1)} &\leftarrow& 2q^{(k+1)} - q^{(k)}\\
      k &\leftarrow& k + 1
    \end{eqnarray*}
  }
  \label{Tab:algo_simplified}
\end{algorithm}

@@

\section{Results}
\paragraph{Simulated game}
See Figure \ref{Tab:sim_dgap_curve}.

@@

%% The framework presented sofar only demanded we be able to efficiently compute metric projections onto the nonnegative orthant $\mathbb{R}^{n_k}_+$, namely nonnegative clipping. This opens perspectives to envisage general two-person zero sum games for which the players' strategy profiles are of the form
%% \begin{equation}
%%   Q_k := \{z \in C_k| E_kz = e_k\text\}
%%   \label{eq:q_gen}
%% \end{equation}
%% where the $C_k$ are arbitrary convex sets for which the metric projections $\Pi_{C_k}$ can be effectively computed.
%% % \paragraph{Example: Computing best response strategies}

@@

%% One observes that this problem is an instance of problem \eqref{eq:primal_pb} with
%% $Q_2 = \{y_0\} = \{z \in C_2| E_2z = e_2\text\}$ and $C_2 = \{y_0\}$. Of course the metric projection on $C_2$ is simply the constant map $y \mapsto y_0$.

>>>>

In \cite{von1996efficient, vonequilibrium}, it was established that for sequential two-person zero-sum games with imcomplete information and perfect recall, there exist sparse matrices $A \in \mathbb{R}^{n_1 \times n_2}, E_1 \in \mathbb{R}^{l_1 \times n_1}, E_2 \in \mathbb{R}^{l_2 \times n_2}$, and vectors $e_1 \in \mathbb{R}^{l_1}, e_2 \in \mathbb{R}^{l_2}$ such that $n_1$, $n_2$, $l_1$, and $l_2$ are all linear in the size of the game tree (number of states in the game) and such that Nash equilibria correspond to pairs $(x, y)$ of \textit{realization plans} which solve the primal LCP (Linear Constrained Programming) problem
\begin{equation}
  \begin{aligned}
    & \underset{y,p}{\text{minimize}}
    & & e_1^Tp \\
    & \text{subject to}
    & & -Ay + E_1^Tp \geq 0,\\
    &&& E_2y = e_2,\\
    &&& y \ge 0
  \end{aligned}
  \label{eq:primal_pb}
\end{equation}

@@

with \textit{duality gap} given by
\begin{equation}
  \mathcal{G}(p, q) := e_1^Tp - (-e_2^Tq) = e_1^Tp + e_2^Tq
  \label{eq:dgap}
\end{equation}

@@

Of course by \textit{weak duality} in problems \eqref{eq:primal_pb} and \eqref{eq:dual_pb}, it holds that
\begin{equation}
  \mathcal{G}(p, q) \geq 0
\end{equation}
for any fixed pair $(x, y)$ of realization plans and admissible dual pair $(p,q)$ in problems \eqref{eq:primal_pb} and \eqref{eq:dual_pb}.

It was also shown (for example, Theorem 3.14 of \cite{vonequilibrium}) that a $(x, y)$ is a solution to the primal-dual problem \eqref{eq:primal_pb} and \eqref{eq:dual_pb} iff there exist vectors $p$ and $q$ such that
\begin{equation}
  \begin{split}
    E_1x = e_1,\text{ }x \geq 0,\hspace{1em} E_2y = e_2,\text{ }y \geq 0\\
    -Ay + E_1^Tp \geq 0,\hspace{1em}-A^Tx + E_2^Tq \leq 0
    \end{split}
\end{equation}

Moreover, in this case \textit{strong duality} holds and the value of the game equals $e_1^Tp = -e_2^Tq$; thus the duality gap $\mathcal{G}(p, q)$ defined in \eqref{eq:dgap} vanishes.

@@

\paragraph{Outline of article}
The goal of this article is to device a simple algorithm which produces a Nash $\epsilon$-equilibrium after $\mathcal{O}(1/\epsilon)$ iterations%We first briefly introduce the sequence-form representation, and contrast it to the strategic-form representation. Then we state the Nash equilibrium problem for the sequence-form representation. Then in
In section \ref{sec:related_work}, we give a brief overview of existing methods for solving \eqref{eq:approx_pb}. In section \ref{sec:algo} we derive the proposed algorithm for solving this problem and prove its $\mathcal{O}(1/\epsilon)$ convergence rate. With construct with a demonstration on Kuhn 3-card Poker.

@@

\section{Related work}
\label{sec:related_work}
In \cite{hoda2010smoothing}, a nested iterative procedure using the Excessive Gap Technique \cite{nesterov2005excessive} was used to to solve the LCP problems \eqref{eq:primal_pb} and \eqref{eq:dual_pb} in the following equivalent saddle-point form

\begin{equation}
  \underset{y \in Q_2}{\text{minimize}}\text{ }\underset{x \in Q_1}{\text{maximize}}\text{ }x^TAy
\end{equation}

They authors reported a $\mathcal{O}(1/\epsilon)$ convergence rate (which derives from the general EGT theory) for the outer-most iteration loop, though the cost of an iteration is presumably heavy since it entails computing a $softmax$ at each iteration.

\section{Useful convex analysis}
\label{sec:notation}
We will need the following notations in the sequel. Given positive integers $m$ and $n$, $\mathbb{R}^{m \times n}$ denotes
the space of all $m$-by-$n$ real matrices. $0_{m,n}$ denotes the $m$-by-$n$ matrix of zeros and $1_{m,n}$ denotes the $m$-by-$n$ matrix of ones.
%% $\mathbb{R}^n_+$ := $\{x \in \mathbb{R}^n|x_k \ge 0 \text{ }  \forall j\}$ is the $n$-dimensional \textit{nonnegative orthant}.

@@

Note that $i_{C \cap D} = i_C + i_D$. The \textit{euclidean projection operator} onto $C$, denoted $\Pi_C$, is the function
$\Pi: \mathbb{R}^n \mapsto C$ which maps a point $x \in \mathbb{R}^n$ to the (necessarily unique) point $\Pi_C(x)$ of $C$ which is closest to $x$. Precisely,
\begin{equation}
  \Pi_C(x) := \underset{c \in C}{argimin}\text{ }\|c - x\|^2
\end{equation}
For example, the euclidean projections onto the nonnegative orthant and the closed unit-disk $\bar{\Delta}$
% := \{x|\|x\| \le 1\}$
are $\Pi_{\mathbb{R}^n_+}(x) \equiv (x)_+$ and $\Pi_{\bar{\Delta}}(z) \equiv \dfrac{z}{max(\|z\|, 1)}$ respectively.

Let $f : \mathbb{R}^n \rightarrow [-\infty, +\infty]$ be a \textit{proper convex lower semi-continous function}
(\textit{p.c.l.s.c} for short). The \textit{convex conjugate} of $f$ is the function $f^*: \mathbb{R}^n \rightarrow [-\infty, +\infty]$ is defined by
\begin{equation}
  x \mapsto f^*(x) := \underset{z \in \mathbb{R}^n}{\text{max}}\text{ }z^Tx - f(x)
\end{equation}

For example $i_{\mathbb{R}^{n}_+}^*(x) \equiv i_{\mathbb{R}^{n}_+}(-x)$; $(a^Tx)^* \equiv i_{\{0\}}(x - a)$; and $(\frac{1}{2}\|x\|^2)^* \equiv \frac{1}{2}\|x\|^2$. One easily verifies that $f^{**} = f$.

@@

\section{The proposed algorithm}
\label{sec:algo}
We now derive our primal-dual algorithm for solving the Nash equilibrium problem \eqref{eq:primal_pb}. The algorithm is obtained from the general primal-dual Algorithm 39 proposed in \cite{chambolle2010}. Though this scheme has recently gained considerable popularity in the signal processing community
%\cite{gramfort-etal:2013a}
, to the best of our knowledge, this is the first time it is being applied to compute Nash equilibria.

It should be noted that the difficulty of applying the general primal-dual Algorithm 39 proposed in \cite{chambolle2010} directly to problem \eqref{eq:primal_pb} lies in the difficulty of effectively computing the euclidean projections $\Pi_{Q_k}$. Theorem \ref{thm:pd} provides a trick to effectively evade this difficulty.

\begin{theorem}
  The Nash equilibrium LCP problems \eqref{eq:primal_pb} and \eqref{eq:dual_pb} can be re-written in the saddle-point convex-concove form

@@

Now, one computes
\begin{equation}
  \begin{split}
    H^*\left(K\begin{bmatrix}y\\p\end{bmatrix}\right) &= \underset{x,q}{\text{max}}\text{ }\begin{bmatrix}x\\q\end{bmatrix}^TK\begin{bmatrix}y\\p\end{bmatrix} - H(x, q) \\
      &= \underset{x, q}{\text{max}}\text{ }x^T(Ay - E_1^Tp) - i_{\mathbb{R}^{n_1}_+}(x) + q^TE_2y - f^Tq\\
      &= \underset{x \geq 0}{\text{max}}\text{ }x^T(Ay - E_1^Tp) + \underset{q}{\text{max}}\text{ }q^T(E_2y - f)\\
      &= i_{\mathbb{R}^{n_1}_+}(-Ay + E_1^Tp) + i_{\{0\}}(E_2y - e_2)
  \end{split}
  \label{eq:b}
\end{equation}
Combining equations \eqref{eq:a} and \eqref{eq:b}, it follows that the LCP problem \eqref{eq:primal_pb} can be written as 
\begin{equation}
  \underset{y,p}{\text{minimize}}\text{ }G(y,p) + H^*\left(K\begin{bmatrix}y\\p\end{bmatrix}\right)
\end{equation}

@@

But these last two equations are simply the primal and dual formulations of the saddle-point problem \eqref{eq:unconstrained_pb}, and we are done.

@@

\begin{algorithm}[H]
  \caption{$\mathcal{O}(1/\epsilon)$ Primal-dual algorithm for finding a Nash $\epsilon$-equilibrium for a sequential two-person zero-sum game with imcomplete information and perfect recall}
  \KwIn{sequence-form specification of the game: $(A, E_1, E_2, e_1, e_2)$, where $A \in \mathbb{R}^{n_1 \times n_2}$,
  $E_1 \in \mathbb{R}^{l_1 \times n_1}$, $E_2 \in \mathbb{R}^{l_2 \times n_2}$, $e_1 \in \mathbb{R}^{l_1}$, $e_2 \in \mathbb{R}^{l_2}$}
  \KwOut{Nash $\epsilon-$equilibrium pair $(x^{(k)}$, $y^{(k)})$ of realization plans}
  \textbf{Precompute}: $\|K\|^2$, where $K$ is constructed as in equations \eqref{eq:unconstrained_pb}. $\|K\|^2$ can be computed via a \textit{power iteration} on $K^TK$, for example.\\
  \textbf{Initialize}:
  $x^{(0)} \in \mathbb{R}^{n_1}$; $\tilde{y^{(0)}}, y^{(0)} \in \mathbb{R}^{n_2}$; $p \in \mathbb{R}^{l_1}$; $q^{(0)} \in \mathbb{R}^{l_2}$; 
  $\tau, \sigma > 0 \text{ s.t. }\tau\sigma \|K\|^2 < 1$ (for example take $\tau = \sigma = .99/\|K\|$); $k = 0$.\\
  \While{
%$\dfrac{\| x^{(k+1)} - x^{(k)}\|^2 + \|v^{(k+1)}- v^{(k)}\|^2}{\sigma} + \dfrac{\|y^{(k+1)}- y^{(k)}\|^2 + \|u^{(k+1)}- u^{(k)}\|^2}{\tau} < \epsilon$
$|e_1^Tp^{(k)} + e_2^Tq^{(k)}| \geq \epsilon$}{
    \begin{eqnarray*}
      x^{(k+1)} &\leftarrow& \left(x^{(k)} + \tau \left(A\tilde{y}^{(k)} - E_1^T\tilde{p}^{(k)}\right)\right)_+\\
      q^{(k+1)} &\leftarrow& q^{(k)} + \tau \left(E_2\tilde{y}^{(k)} - e_2\right)\\
      y^{(k+1)} &\leftarrow& \left(y^{(k)} - \sigma \left(A^Tx^{(k + 1)} + E_2^Tq^{(k + 1)}\right)\right)_+\\
      p^{(k+1)} &\leftarrow& p^{(k)} - \sigma \left(e_1 - E_1x^{(k+1)}\right)\\
      \tilde{y}^{(k+1)} &\leftarrow& 2y^{(k+1)} - y^{(k)}\\
      \tilde{q}^{(k+1)} &\leftarrow& 2q^{(k+1)} - q^{(k)}\\
      k &\leftarrow& k + 1
    \end{eqnarray*}
  }
  \label{Tab:algo_simplified}
\end{algorithm}

@@

\section{Results}
\paragraph{Simulated game: matrix game on simplexes}
Any matrix $A \in \mathbb{R}^{n_1 \times n_2}$ specifies a 
As in \textbf{nesterov, c-p, etc.}, we generate a $1000 \times 1000$ random matrix whose entries are uniformly identically distributed in the closed interval $[-1, 1]$.
See Figure \ref{Tab:sim_dgap_curve}.

@@

%% The framework presented sofar only demanded we be able to efficiently compute euclidean projections onto the nonnegative orthant $\mathbb{R}^{n_k}_+$, namely nonnegative clipping. This opens perspectives to envisage general two-person zero sum games for which the players' strategy profiles are of the form
%% \begin{equation}
%%   Q_k := \{z \in C_k| E_kz = e_k\text\}
%%   \label{eq:q_gen}
%% \end{equation}
%% where the $C_k$ are arbitrary convex sets for which the euclidean projections $\Pi_{C_k}$ can be effectively computed.
%% % \paragraph{Example: Computing best response strategies}

@@

%% One observes that this problem is an instance of problem \eqref{eq:primal_pb} with
%% $Q_2 = \{y_0\} = \{z \in C_2| E_2z = e_2\text\}$ and $C_2 = \{y_0\}$. Of course the euclidean projection on $C_2$ is simply the constant map $y \mapsto y_0$.
